---
title: "Bayesian Network Model & Counterfactuals with binned data"
output: html_notebook
---

```{r}
# libraries
library(bnlearn) # bayesian networks library
library(dplyr)
library(ggplot2)
library(tidyr)
```


# The Data
```{r}
path_2_df <- "/Users/masinde/Projects/causal-fairness-Philippines-drrm/data/mun_data.RData"

load(file = path_2_df)

head(df)
```

## Correct column name
```{r}
# correct column name yellow_ss_frac2 to yellow_ls_frac
# GEN_Yellow_per_LSbl
clnd_municipality_tc_df <- clnd_municipality_tc_df %>%
  rename(yellow_ls_frac = yellow_ss_frac2)
```


## Removing not needed columns
```{r}
# Remove "name" and "age" columns
df <- clnd_municipality_tc_df %>% select(-housing_units, -pantawid_benef, 
                            -Population.2020.Census., -Mun_Code_2, -Key)

head(df)
```
## Removing NA's
```{r}
# remove the NA's
df <- df %>% 
  drop_na()

nrow(df)
```


## Column datatypes
```{r}
# Get column data types using base R
column_types <- sapply(df, class)

print(column_types)
```


Change variables "has_coast", "Income.Class", Mun_Code_3 to factor
```{r}
# Convert specific variables to factors
df <- df %>%
  mutate(
    has_coast = factor(has_coast),
    Income.Class = factor(Income.Class),
    Mun_Code_3 = factor(Mun_Code_3)
  )

# Check the result
str(df)
```


# Bayesian Network Modeling

## Test and Train set
```{r}
# split into train, test set
# n as number of observations
n <- nrow(df)

# for reproducibility
set.seed(12345)

# use 40, 30, 30  split
id <- sample(1:n, floor(n*0.6))
df_train <- df[id, ]
df_test <- df[-id, ]
```


```{r}
nrow(df_train)
```

```{r}
nrow(df_test)
```


### Removing extra columns
We do not need the municipality names codes and regions for now. We can merge
later.

Also removing rain_max6h and rain_ma24h which are highly correlated with rain_total
```{r}
bn_train <- df_train %>%
  select(-Mun_Code, -Municipality,- Mun_Code_3, -typhoon, -region, -island_groups,
         -vulnerable_groups, -rain_max6h, -rain_max24h, -poverty_pct)

bn_test <- df_test %>%
  select(-Mun_Code, -Municipality, -Mun_Code_3, -typhoon, -region, -island_groups,
         -vulnerable_groups, -rain_max6h, -rain_max24h, -poverty_pct)
```


## The DAG
```{r}
# start with an empty graph
col_names <- colnames(bn_train)
graph_1 <- empty.graph(col_names)
```

```{r}
# setting arcs to the nodes
model_1_arcs <-  matrix(c( # closeness to tropical cyclone
              "track_min_dist", "DAM_perc_dmg",
              "wind_max","DAM_perc_dmg", # direct effect of wind on damage
              "rain_total", "DAM_perc_dmg",
              "ls_risk_pct","DAM_perc_dmg",# number of houses in landslide risk zones
              "ss_risk_pct","DAM_perc_dmg", # number of houses in storm surge risk zones
              "roof_strong_wall_strong", "DAM_perc_dmg",
              "roof_strong_wall_light", "DAM_perc_dmg",
              "roof_strong_wall_salv","DAM_perc_dmg",
              "roof_light_wall_strong","DAM_perc_dmg",
              "roof_light_wall_light","DAM_perc_dmg",
              "roof_light_wall_salv","DAM_perc_dmg",
              "roof_salv_wall_strong","DAM_perc_dmg",
              "roof_salv_wall_light","DAM_perc_dmg",
              "roof_salv_wall_salv","DAM_perc_dmg",
              "has_coast", "coast_length",
              "coast_length","blue_ss_frac",
              "coast_length","yellow_ss_frac",
              "coast_length", "orange_ss_frac",
              "coast_length", "red_ss_frac",
              "blue_ss_frac", "DAM_perc_dmg",
              "yellow_ss_frac", "DAM_perc_dmg",
              "orange_ss_frac", "DAM_perc_dmg",
              "red_ss_frac", "DAM_perc_dmg",
              "elev_mean", "DAM_perc_dmg",
              "ruggedness_mean", "DAM_perc_dmg",
              "slope_mean", "DAM_perc_dmg"), 
            ncol = 2, byrow = TRUE, 
            dimnames = list(NULL, c("from", "to"))
            )
```

```{r}
arcs(graph_1) <- model_1_arcs

```


## Model Fitting and Testing
```{r}
# train model
fit_model_1 <- bn.fit(graph_1, data = bn_train)

# test model on test set
preds_model_1 <- predict(fit_model_1, node = "DAM_perc_dmg", data = bn_test)

# evaluation metrics
MSE_model_1 <- (sum((bn_test$DAM_perc_dmg - preds_model_1)^2)/length(preds_model_1))

MSE_model_1
```

```{r}
# RMSE
sqrt(MSE_model_1)
```

