---
title: "R Notebook"
output: html_notebook
---

# DATA CLEANING

## pacakges
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
library(readxl)
```

## Importing Data
```{r}
path_2_base_df <- "/Users/masinde/Projects/causal-fairness-Philippines-drrm/data/base_inc_data.csv"

base_df <- read.csv(file = path_2_base_df, header = TRUE)

head(base_df)
```

```{r}
path_2_combined_df <- "/Users/masinde/Projects/causal-fairness-Philippines-drrm/data/combined_input_data.csv"

combined_df <- read.csv(file = path_2_combined_df, header = T)

head(combined_df)
```


## Renaming the columns
From here on we work with the "base_inc_data". We rename it to "municipality_cyclone_df"
```{r}
# renaming the columns
municipality_tc_df <- base_df %>%
  rename(
    rain_total = HAZ_rainfall_Total,
    rain_max6h = HAZ_rainfall_max_6h,
    rain_max24h = HAZ_rainfall_max_24h,
    wind_max = HAZ_v_max,
    track_min_dist = HAZ_dis_track_min,
    ls_risk_pct = GEN_landslide_per,
    ss_risk_pct = GEN_stormsurge_per,
    blue_ss_frac = GEN_Bu_p_inSSA,
    blue_ls_frac = GEN_Bu_p_LS,
    red_ls_frac = GEN_Red_per_LSbldg,
    orange_ls_frac = GEN_Or_per_LSblg,
    yellow_ss_frac = GEN_Yel_per_LSSAb,
    red_ss_frac = GEN_RED_per_SSAbldg,
    orange_ss_frac = GEN_OR_per_SSAbldg,
    yellow_ss_frac2 = GEN_Yellow_per_LSbl,
    slope_mean = TOP_mean_slope,
    elev_mean = TOP_mean_elevation_m,
    ruggedness_sd = TOP_ruggedness_stdev,
    ruggedness_mean = TOP_mean_ruggedness,
    slope_sd = TOP_slope_stdev,
    has_coast = GEN_with_coast,
    coast_length = GEN_coast_length,
    poverty_pct = VUL_poverty_perc,
    housing_units = VUL_Housing_Units,
    roof_strong_wall_strong = VUL_StrongRoof_StrongWall,
    roof_strong_wall_light = VUL_StrongRoof_LightWall,
    roof_strong_wall_salv = VUL_StrongRoof_SalvageWall,
    roof_light_wall_strong = VUL_LightRoof_StrongWall,
    roof_light_wall_light = VUL_LightRoof_LightWall,
    roof_light_wall_salv = VUL_LightRoof_SalvageWall,
    roof_salv_wall_strong = VUL_SalvagedRoof_StrongWall,
    roof_salv_wall_light = VUL_SalvagedRoof_LightWall,
    roof_salv_wall_salv = VUL_SalvagedRoof_SalvageWall,
    vulnerable_groups = VUL_vulnerable_groups,
    pantawid_benef = VUL_pantawid_pamilya_beneficiary
  )
```


```{r}
head(municipality_tc_df)
```

## Adding more variables

## Removing NA's
We care about NA's in the outcome variable. So we remove rows with NA's in the 
outcome varaible first.
```{r}
# removing NA's rows in DAM_perc_dmg
municipality_tc_df <- municipality_tc_df  %>% 
  filter(!is.na(DAM_perc_dmg))
```


```{r}
nrow(municipality_tc_df)
```

```{r}
head(municipality_tc_df)
```

### Missing municipality names
```{r}
# checking which correspondence code are missing
missing_mun_names <- municipality_tc_df %>% 
  filter(is.na(Correspondence.Code)) %>%
  select(Mun_Code, Mun_Code_2)

nrow(missing_mun_names)
```

```{r}
length(unique(missing_mun_names$Mun_Code))
```

Add the missing municipality names:
```{r warning=FALSE}
# From Philippines Standard Geographic Code (psgc) file
path_2_psgc <- "/Users/masinde/Projects/causal-fairness-Philippines-drrm/data/PSGC-2Q-2024-Publication-Datafile-rev.xlsx"

col_types = c("numeric", "text", "text", "text", "text", "text", "text", "text", "text", "numeric",
              "text")
psgc_2q_2024 <- read_excel(path = path_2_psgc, sheet = "PSGC", col_types = col_types) 

head(psgc_2q_2024)
```

```{r}
# Codes of the missing municipalities
missing_mun <- unique((missing_mun_names$Mun_Code_2))

# convert to character and append 0 if the length is less than 9
cor_missing_mun <- ifelse(nchar(as.character(missing_mun)) < 9,
                 paste0("0", as.character(missing_mun)),  # Append zero
                 as.character(missing_mun))              # Otherwise, keep as is

cor_missing_mun                     
```

```{r}
missing_mun_names <- psgc_2q_2024 %>% 
  filter(`Correspondence Code` %in% cor_missing_mun) %>% # match character
  select(`Correspondence Code`, Name)

missing_mun_names
```

```{r}
# adding the mun names to the dataframe

clnd_municipality_tc_df <- municipality_tc_df %>% 
  mutate(Municipality = case_when(
    Mun_Code_2 %in% as.integer(missing_mun_names$`Correspondence Code`) ~ 
      missing_mun_names$Name[match(Mun_Code_2, as.integer(missing_mun_names$`Correspondence Code`))],
    TRUE ~ Municipality
  ))


```

```{r}
# checking again which municipality names are missing:
any((is.na(clnd_municipality_tc_df$Municipality)))
```

```{r}
is.null(clnd_municipality_tc_df$Municipality)
```

# Drop duplicate columns:
```{r}
clnd_municipality_tc_df <- clnd_municipality_tc_df %>%
  select(-c("X10.Digit.Code","Unnamed..0", "Correspondence.Code"))
```

# Add Regions:
Regional data (corresponding code and names) are obtained from the statistics office website: https://psa.gov.ph/classification/psgc. Using the **way-back machine** for a snapshot of the year *19th June 2020*. Particularly the file named *PSGC Publication March2020.xlsx*. The file is small so we enter the data manually.

Island groups we get from Wikipedia: https://en.wikipedia.org/wiki/Island_groups_of_the_Philippines

WE DO NOT HAVE THE NIR REGION IN THE DATASET

```{r}
# creating a dataframe with regions:
region <- c('National Capital Region', 'Cordillera Administrative Region', 
               'Ilocos Region', 'Cagayan Valley', 'Central Luzon',
               'CALABARZON', 'MIMAROPA Region',
              'Bicol Region', 'Western Visayas',
              'Central Visayas', 
               'Eastern Visayas', 'Zamboanga Peninsula',
              'Northern Mindanao', 'Davao Region', 
               'SOCCSKSARGEN', 'Caraga', 
               'ARMM')

region_code <- c('1300000000', '1400000000', '0100000000', '0200000000', '0300000000', '0400000000', 
             '1700000000', '0500000000', '0600000000', '0700000000',
            '0800000000', '0900000000', '1000000000', '1100000000', '1200000000',
            '1600000000', '1500000000')


island_groups <- c("Luzon", "Luzon", "Luzon", "Luzon", "Luzon", "Luzon", "Luzon", "Luzon",
                   "Visayas", "Visayas", "Visayas", "Mindanao", "Mindanao", "Mindanao",
                   "Mindanao", "Mindanao", "Mindanao"
                   )

regions_df <- as.data.frame(cbind(region_code, region, island_groups))

head(regions_df)
```

```{r}
# create mun_code 3 which is character and correct length
clnd_municipality_tc_df$Mun_Code_3 <- ifelse(nchar(as.character(clnd_municipality_tc_df$Mun_Code_2)) < 9,
                 paste0("0", as.character(clnd_municipality_tc_df$Mun_Code_2)),  # Append zero
                 as.character(clnd_municipality_tc_df$Mun_Code_2))              # Otherwise, keep as is
```


```{r}
# merge regions_df with cln_municipality_tc_df

clnd_municipality_tc_df <- clnd_municipality_tc_df %>%
  # Extract the first two characters of D
  mutate(Key = substr(Mun_Code_3, 1, 2)) %>%
  # Perform a left join with B
  left_join(
    regions_df %>%
      mutate(Key = substr(region_code, 1, 2)) %>% 
      select(Key, region, island_groups), # Keep only the relevant columns from B
    by = "Key"
  ) #%>%
  # Rename F to E (if desired)
  # rename(E = F) %>%
  # Drop the intermediate Key column
  #select(-Key)

head(clnd_municipality_tc_df)
```

```{r}
# any missing?
any(is.na(clnd_municipality_tc_df$region))
```


```{r}
# add island groups
# merge regions_df with cln_municipality_tc_df

clnd_municipality_tc_df <- clnd_municipality_tc_df %>%
  # Extract the first two characters of D
  #mutate(Key = substr(Mun_Code_3, 1, 2)) %>%
  # Perform a left join with B
  left_join(
    regions_df %>%
      mutate(Key = substr(region_code, 1, 2)) %>% 
      select(Key, region), # Keep only the relevant columns from B
    by = "Key"
  ) #%>%
  # Rename F to E (if desired)
  # rename(E = F) %>%
  # Drop the intermediate Key column
  #select(-Key)

head(clnd_municipality_tc_df)
```


```{r}
# Add the three regions
```


# Export data
```{r}
# path to write cleaned data 
path_2_write <- "/Users/masinde/Projects/causal-fairness-Philippines-drrm/data"
write.csv(clnd_municipality_tc_df, file = paste(path_2_write, "mun_model_data.csv", sep = "/"))
```



# Data documentation
```{r}
# Create metadata as a list
metadata <- list(
  rain_total = list(type = "numeric", 
                    description = "Total rainfall (mm)", 
                    constraints = ">= 0"),
  rain_max6h = list(type = "numeric",
                    description = "Maximum rainfall within a 6h period (mm)", 
                    constraints = ">= 0"),
  rain_max24h = list(type = "numeric",
                    description = "Maximum rainfall within a 24h period (mm)", 
                    constraints = ">= 0"),
  wind_max = list(type = "numeric", 
                  description = "Max 1 min sustained wind speed, based on wind field (ms−1)", 
                  constraints = ">= 0"),
  track_min_dist = list(type = "numeric", 
                  description = "Minimum distance between typhoon track and  municipality", 
                  constraints = ">= 0"),
  ls_risk_pct = list(type = "numeric", 
                  description = "Percentage of houses (OSM footprint) in landslide risk zones", 
                  constraints = ">= 0"),
  ss_risk_pct = list(type = "numeric", 
                  description = "Percentage of houses (OSM footprint) in storm surge risk zones", 
                  constraints = ">= 0"),
  blue_ss_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored blue in storm surge risk map", 
                  constraints = ">= 0"),
  blue_ls_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored blue in landslide risk map", 
                  constraints = ">= 0"),
  red_ls_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored red in landslide risk map", 
                  constraints = ">= 0"),
  orange_ls_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored orange in landslide risk map", 
                  constraints = ">= 0"),
  yellow_ss_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored yellow in storm surge risk map", 
                  constraints = ">= 0"),
  red_ss_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored red in storm surge risk map", 
                  constraints = ">= 0"),
  orange_ss_frac = list(type = "numeric", 
                  description = "Fraction of municipality colored orange in storm surge risk map", 
                  constraints = ">= 0"),
  yellow_ss_frac2 = list(type = "numeric", 
                  description = "Fraction of municipality colored yellow in storm surge risk map", 
                  constraints = ">= 0"),
  slope_mean = list(type = "numeric", 
                  description = "Slope mean", 
                  constraints = ">= 0"),
  elev_mean = list(type = "numeric", 
                   description = "Elevation mean", 
                   constraints = ">= 0"),
  ruggedness_sd = list(type = "numeric", 
                       description = "Ruggedness standard deviation", 
                       constraints = "0-100"),
  ruggedness_mean = list(type = "integer", 
                         description = "Ruggedness mean", 
                         constraints = ">= 0"),
  slope_sd = list(type = "integer", 
                         description = "Slope standard deviation", 
                         constraints = ">= 0"),
  poverty_pct = list(type = "integer", 
                         description = "Percentage of people in poverty", 
                         constraints = ">= 0"),
  has_coast = list(type = "categorical", 
                         description = "Boolean: coast or no coast", 
                         constraints = ">= 0"),
  coast_length = list(type = "integer", 
                         description = "Length of coast", 
                         constraints = ">= 0"),
  housing_units = list(type = "integer", 
                         description = "Total number of housing units", 
                         constraints = ">= 0"),
  roof_strong_wall_strong = list(type = "integer", 
                         description = "Number of houses with a strong roof and strong walls", 
                         constraints = ">= 0"),
  roof_strong_wall_light = list(type = "integer", 
                         description = "Number of houses with a strong roof and light walls", 
                         constraints = ">= 0"),
  roof_strong_wall_salv = list(type = "integer", 
                         description = "Number of houses with a strong roof and salvaged walls", 
                         constraints = ">= 0"),
  roof_light_wall_strong = list(type = "integer", 
                         description = "Number of houses with a light roof and strong walls", 
                         constraints = ">= 0"),
  roof_light_wall_light = list(type = "integer", 
                         description = "Number of houses with a light roof and light walls", 
                         constraints = ">= 0"),
  roof_light_wall_salv = list(type = "integer", 
                         description = "Number of houses with a light roof and salvaged walls", 
                         constraints = ">= 0"),
  roof_salv_wall_strong = list(type = "integer", 
                         description = "Number of houses with a salvaged roof and strong walls", 
                         constraints = ">= 0"),
  roof_salv_wall_light = list(type = "integer", 
                         description = "Number of houses with a salvaged roof and light walls", 
                         constraints = ">= 0"),
  roof_salv_wall_salv = list(type = "integer", 
                         description = "Number of houses with a salvaged roof and salvaged walls", 
                         constraints = ">= 0"),
  vulnerable_groups = list(type = "integer", 
                         description = "Vulnerable groups from DSWD National Household Targeting Office", 
                         constraints = ">= 0"),
  pantawid_benef = list(type = "integer", 
                         description = "Number of Pantawid Pamilya beneficiary households", 
                         constraints = ">= 0")
)
```

# 
